\documentclass[12pt,a4paper]{article}

\usepackage[margin=1in]{geometry}
\usepackage[english]{babel}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

\usepackage{longtable}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage[parfill]{parskip}
\usepackage[title,titletoc]{appendix}
\usepackage{amsmath}
\usepackage{fnpos}
\usepackage{geometry}
\usepackage{csquotes}
\usepackage{tikz}
\usepackage{pdflscape}

\usetikzlibrary{positioning}
\usetikzlibrary{calc}
 \tikzstyle{main} = [rectangle, rounded corners, minimum height=0.7cm,text centered, draw=black, font=\scriptsize]
 \tikzstyle{process} = [rectangle, minimum height=0.7cm, text centered, draw=black, font=\scriptsize]
 \tikzstyle{note} = [rectangle, text centered, minimum height=1cm]
 \tikzstyle{arrow} = [->,>=stealth,auto,font=\scriptsize,node distance=3cm, main node/.style={circle,draw}]

\usepackage[hyphens]{url}

\usepackage[backend=biber,url=false,style=authoryear,sorting=nyt,bibstyle=numeric]{biblatex}
\addbibresource{chequeBounce.bib}

\usepackage[acronym,nonumberlist,nomain]{glossaries}
\makeglossary
\loadglsentries{acronyms.tex}

\AtBeginEnvironment{quote}{\smaller}
\makeFNbottom
\geometry{margin=1in}
\renewcommand{\baselinestretch}{1.25}
\setlength{\parskip}{1em}
\setlist{nosep}

\newcommand{\floattabu}[1]{
\vspace*{0.2in}
{\footnotesize
#1
}
\vspace*{0.2in}}

\usepackage[hidelinks,breaklinks]{hyperref}
\usepackage{cleveref}

\title{Characterising cheque dishonour cases in India: Causes for delays and policy implications}
\author{Devendra Damle\thanks{Devendra Damle is a researcher at the National Institute of Public Finance and Policy.} and Jitender Madaan\thanks{Jitender Madaan is an Associate Professor at the Indian Institute of Technology, Delhi.} and Karan Gulati\thanks{Karan Gulati is a researcher at the Vidhi Centre for Legal Policy.}\\ and Manish Kumar Singh\thanks{Manish Kumar Singh is an Assistant Professor at the Indian Institute of Technology, Roorkee.} and Nikhil Borwankar\thanks{Nikhil Borwankar is a practicing advocate.}}

\begin{document}
\maketitle

\begin{abstract}

This paper analyses the determinants of case disposition time under the Negotiable Instruments Act, 1881 by creating a unique dataset of roughly 50,000 case-level records spread across ten major States in India. Unlike previous studies, our analysis focuses mainly on procedural matters and specific case-characteristics to estimate the case duration. Regression results show that multiplicity of proceedings, non-appearance of accused, trials being run as summons trial, other jurisdictional issues, and reference to mediation significantly increases the case duration. These findings are robust even when we consider number of hearings as a proxy for case disposition time. Additionally, results indicate that cases being contested show lower case duration though it takes three additional hearings to resolve. Our estimates also indicate that 2015 amendment focused on clarifying the jurisdiction and mandatory centralisation of cases against the same drawer has reduced the case duration by six month within a span of two years.

%Our results also indicate that 2015 amendments focused on clarifying the jurisdiction related issues for filing cases and mandatory centralisation of cases against the same drawer has led to a major decrease of case duration. Our estimate suggest that it has reduced the case duration on average by more than six months within a span of two years.
		
	%Regression results show that cases referred for mediation on average takes 108 additional days to resolve. Additionally, results indicate that multiplicity of proceedings, non-appearance of accused, trials being run as summons trial, other jurisdictional issues 
		
	%The findings are robust when we consider number of hearings as a proxy for case disposition time. We find 3.27 additional hearing for cases sent for mediation.
		
	%Furthermore, if we consider number of hearings as a proxy for case disposition time, we find 3.27 additional hearing for cases sent for mediation.

\end{abstract}

\textit{Keywords: } Negotiable Instruments Act, cheque bouncing, disposition time, court delay, procedural reforms

\textit{JEL Classification:} K12, K19, K41  

\newpage
\tableofcontents

\newpage
\printglossaries

\newpage
\listoftables

\newpage
\section{Introduction}
\label{sec:introduction}
\input{introduction}

\section{Methodology}
\label{sec:methodology}
\input{methodology}

\section{Results}
\label{sec:results}
\input{results}

\section{Conclusion}
\label{sec:conclusion}
\input{conclusion}

\section*{Acknowledgements}
The authors would like to thank the DAKSH Centre of Excellence for Law and Technology for their support. The authors are grateful to thank Sandhya PR and Surya Prakash BS for their extensive feedback on the study. The authors would also like to thank Manaswini Rao and Pramod Rao for reviewing the study and their inputs. We also thank Abhishek Seth for his excellent research assistance. All errors are ours.

\section*{Funding}
This work was supported by the DAKSH Centre of Excellence (CoE) for Law and Technology at IIT Delhi.

\newpage
\printbibliography[heading=bibintoc]

\newpage
\begin{appendices}
\input{appendices}
\end{appendices}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
